---
title: "Date Shift Method"
author: "Veronica I. Underwood Carrasco & Katrina Ramsey"
format: html
editor: visual
---

# Data Shift in R

## Consider this example dataset

```{r, include=FALSE}
origchildtbl <- data.frame(Study_ID = c("0000", 6834, 4142, 4290, 6563, 2906)
               , Original_Date_1 = c("1/25/2016", "3/24/2017", "4/30/2017", "1/17/2016",  "8/3/2018", "6/17/2016")
               , Original_Date_2 = c("9/23/2016", "8/6/2018",  "1/30/2018", "3/2/2016",  "9/17/2018", "9/24/2016"))
```

```{r}
origchildtbl
```

### Make sure date columns can be read as dates

```{r}
#| warning: FALSE
library(lubridate) #package needed to set date format

#read dates as date format
origchildtbl$Original_Date_1 <- lubridate::mdy(origchildtbl$Original_Date_1)
origchildtbl$Original_Date_2 <- lubridate::mdy(origchildtbl$Original_Date_2)

origchildtbl
```

## Create data folders for saving final results

```{r}
#| warning: FALSE
# Best practice is to create a Project for your work
library(here) #package needed to auto-locate your current working directory
studyfld <- here()

# create separate folders for crosswalk table and final de-identified data
crosswalkfld <- paste(studyfld, "Crosswalk",     #naming location
                      sep = "/")
deidentfld   <- paste(studyfld, "Deidentified",  #naming location
                      sep = "/")
```

## Initial steps for reproducibility

```{r}
# set random number seed so that commands will produce the same set of random numbers each time it is run
set.seed(1234321) #example

#sort the original table by unique study identifier to make sure you can reproduce results in the same way
newchildtbl <- origchildtbl[order(origchildtbl$Study_ID),]

newchildtbl
```

### Note. Converting from Child ID to Study ID

```{r}
#| warning: false
library(dplyr)    #for data transformation
library(magrittr) #for pipe operator

## if starting with system-specific child ID (i.e., an identifier)
newchildtbl1 <- newchildtbl %>% 
                     rename(Child_ID = Study_ID) #changing label for example

newchildtbl1

# generate a uniform random number and sort by it
newchildtbl1$rand1 <- runif(n = nrow(newchildtbl1))
newchildtbl1 <- newchildtbl1[order(newchildtbl1$rand1),]

# then add an ID that is the row number in the new, random order
newchildtbl2 <- dplyr::mutate(newchildtbl1, Study_ID = row_number())
#drop rand1 column 
newchildtbl2 <- newchildtbl2 %>%
                select(Child_ID, Original_Date_1, Original_Date_2, Study_ID) 

newchildtbl2
```

## Date Shift

```{r}
#new column for date shift integer
newchildtbl$dateshift <- sample(-21:21              # the range
                                , nrow(newchildtbl) # number of rows to generate
                                , replace=T)        # with replacement (allow numbers to be reused)

newchildtbl
```

### Apply Date Shift to Original Dates

```{r}
newchildtbl$New_Date_1 <- newchildtbl$Original_Date_1 + newchildtbl$dateshift
newchildtbl$New_Date_2 <- newchildtbl$Original_Date_2 + newchildtbl$dateshift
```

### Reorder and select needed columns

```{r}
newchildtbl <- newchildtbl %>%
                    select(Study_ID, Original_Date_1, Original_Date_2,
                           dateshift, New_Date_1, New_Date_2)

newchildtbl
```

#### If starting with Child ID:

```{r, include=FALSE}
#new column for date shift integer
newchildtbl2$dateshift <- sample(-21:21              # the range
                                , nrow(newchildtbl2) # number of rows to generate
                                , replace=T)         # with replacement (allow numbers to be reused)

#apply date shift
newchildtbl2$New_Date_1 <- newchildtbl2$Original_Date_1 + newchildtbl2$dateshift
newchildtbl2$New_Date_2 <- newchildtbl2$Original_Date_2 + newchildtbl2$dateshift

#reorder and select needed columns 
newchildtbl2 <- newchildtbl2 %>%
                    select(Child_ID, Original_Date_1, Original_Date_2,
                           dateshift, Study_ID, New_Date_1, New_Date_2)
```

```{r}
newchildtbl2
```

## Save two files

### Cross-walk copy for future reference

```{r}
#all variables, including identifiers
xwalkfn <- paste(crosswalkfld, "CrosswalkChildTable.RDS", sep = "/")
saveRDS(newchildtbl2, file = xwalkfn)
```

### De-identified copy for sharing with collaborating researchers

```{r}
# study-specific, de-identified variables 
deident <- select(newchildtbl, Study_ID, New_Date_1, New_Date_2) # select columns
deidentfn <- paste(deidentfld, "DeIDChildTable.csv", sep = "/")  # file name and path (Windows)
write.csv(deident            # dataframe name to export (csv for general use)
          , file = deidentfn # to the file and path defined above
          , row.names = FALSE)
```
